---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)

suppressPackageStartupMessages(library(magrittr, quietly = TRUE))
suppressPackageStartupMessages(library(pmatch, quietly = TRUE))

```

# pmatch -- A package for Haskell-like pattern matching in R

The goal of the `pmatch` package is to provide structure pattern matching, similar to Haskell and ML, to R programmers. The package provide functionality for defining new types and for matching against the structure of such types.

The idea behind pattern matching is that we define types by how we create them, and we have ways of matching a pattern of constructors against a value to pick the one that matches the value.

The simplest example is a type defined just from constants. For example, we can define the type `enum` to consist of one of `ONE`, `TWO`, or `THREE`.

```{r}
enum := ONE | TWO | THREE
```

Any of these three constants will be created by this command. If you print them, they just give you their names:

```{r}
ONE
TWO
THREE
```

The interesting feature is that we can match against these constructor-constants. Using the `cases` function we can pick a pattern that matches a value.

```{r}
elements <- list(ONE, TWO, THREE)
for (elm in elements) {
    value <- cases(elm,
                   ONE -> 1,
                   TWO -> 2,
                   THREE -> 3)
    cat("Element", toString(elm), "maps to value", toString(value), "\n")
}
```

The `cases` function works by matching its first element--which should be a value constructed as a type we have defined with `:=`--against a list of patterns and what `cases` should return if a pattern matches. The pattern arguments are on the form `pattern -> expression`. The value is matched against the patterns in turn and the first pattern that matches will be chosen. The expression to the right of the arrow is then evaluated and the result is returned.

The patterns do not need to be literal constants. You can also use variables. These will be bound to the matching value and the expression that is evaluated will see such variables bound.

```{r}
elements <- list(ONE, TWO, THREE)
for (elm in elements) {
    value <- cases(elm,
                   ONE -> 1,
                   v -> v)
    cat("Element", toString(elm), "maps to value", toString(value), "\n")
}
```

It gets more interesting when we move beyond constants. The `:=` operator also allows you to define function-constructors. These are written simply as you would write a function call, but the variables are interpreted as parameters of the constructor. For example, we could define:

```{r}
zero_one_two_three := ZERO | ONE(x) | TWO(x,y) | THREE(x,y,z)
```

The first constructor, `ZERO`, is just a constant as before, but the other three takes arguments.

```{r}
ONE(1)
TWO(1,2)
THREE(1,2,3)
```

When we use `cases` to match against such patterns, we can bind variables to the values they contain.

```{r}
f <- function(v) {
    cases(v,
          ZERO         -> 0,
          ONE(x)       -> x,
          TWO(x,y)     -> x + y,
          THREE(x,y,z) -> x + y + z)
}

f(ZERO)
f(ONE(1))
f(TWO(1,2))
f(THREE(1,2,3))
```

You can nest these patterns to match on more complex values

```{r}
f <- function(v) {
    cases(v,
          ZERO               -> 0,
          ONE(x)             -> x,
          TWO(ONE(x),ONE(y)) -> x + y + 42,
          TWO(x,y)           -> x + y,
          THREE(x,y,z)       -> x + y + z)
}

f(TWO(ONE(10),ONE(-10)))
```

You have to be careful with the order of expressions, though. If we flipped the two `TWO` patterns, the first one, `TWO(x,y)` would match first and we would be trying to add together `ONE(10)` and `ONE(-10)`, which would result in an error since we do not have an addition operator defined on these types.

You can match any pattern using the bare keyword `otherwise`:

```{r}
cases(42,
      1 -> 1,
      13 -> 13,
      otherwise -> 24)
```

When you define function constructors, you can give the arguments types. You do this by adding `:` and a type name to the argument. For example, we could define

```{r}
one_or_two := ONE(x : numeric) | TWO(x : numeric, y : numeric)
```

We would now get an error if the arguments we provide to the constructors were not `numeric`:

```{r, error=TRUE}
ONE(1)
ONE("foo")
```

Constructors and pattern matching becomes even more powerful when you start to define recursive data structures. You can, for example, define a binary tree like this:

```{r}
tree := L(elm : numeric) | T(left : tree, right : tree)
```

A very succinct depth first traversal that collects the leaves of such a tree can be written like this:

```{r}
f <- function(x) {
    cases(x, 
          L(v) -> v, 
          T(left,right) -> c(f(left), f(right)))
}
x <- T(T(L(1),L(2)), T(T(L(3),L(4)),L(5)))
f(x)
```

For more examples, see below.

## Installation

You can install pmatch from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("mailund/pmatch")
```

## Examples

To run the examples below, you will need to use the `magrittr` package for the pipe operator, `%>%`.

```r
library(magrittr)
```


### Linked lists


```{r list-definition}
linked_list := NIL | CONS(car, cdr : linked_list)

lst <- CONS(1, CONS(2, CONS(3, NIL)))
```

```{r list-functions}
reverse_list <- function(lst, acc = NIL) {
  force(acc)
  cases(lst,
        NIL -> acc,
        CONS(car, cdr) -> reverse_list(cdr, CONS(car, acc)))
}

list_length <- function(lst, acc = 0) {
  force(acc)
  cases(lst,
        NIL -> acc,
        CONS(car, cdr) -> list_length(cdr, acc + 1))
}

list_to_vector <- function(lst) {
  n <- list_length(lst)
  v <- vector("list", length = n)
  f <- function(lst, i) {
    force(i)
    cases(lst,
          NIL -> NULL,
          CONS(car, cdr) -> {
            v[[i]] <<- car
            f(cdr, i + 1)
            }
          )
  }
  f(lst, 1)
  v %>% unlist
}

vector_to_list <- function(vec) {
  lst <- NIL
  for (i in seq_along(vec)) {
    lst <- CONS(vec[[i]], lst)
  }
  reverse_list(lst)
}


lst <- vector_to_list(1:5)
list_length(lst)
list_to_vector(lst)
lst %>% reverse_list %>% list_to_vector
```

### Search trees

```{r}
search_tree := E | T(left : search_tree, value, right : search_tree)

insert <- function(tree, x) {
  cases(tree,
        E -> T(E, x, E),
        T(left, val, right) ->
          if (x < val)
            T(insert(left, x), val, right)
          else if (x > val)
            T(left, val, insert(right, x))
          else
            T(left, x, right)
        )
}

member <- function(tree, x) {
  cases(tree,
        E -> FALSE,
        T(left, val, right) -> {
          if (x < val) member(left, x)
          else if (x > val) member(right, x)
          else TRUE
        })
}

tree <- E
for (i in sample(2:4))
  tree <- insert(tree, i)

for (i in 1:6) {
  cat(i, " : ", member(tree, i), "\n")
}
```

### Red-black search trees

```{r}
colour := R | B
rb_tree := E | T(col : colour, left : rb_tree, value, right : rb_tree)


member <- function(tree, x) {
  cases(tree,
        E -> FALSE,
        T(col, left, val, right) -> {
          if (x < val) member(left, x)
          else if (x > val) member(right, x)
          else TRUE
        })
}

tree <- T(R, E, 2, T(B, E, 5, E))
for (i in 1:6) {
  cat(i, " : ", member(tree, i), "\n")
}

insert_rec <- function(tree, x) {
  cases(tree,
        E -> T(R, E, x, E),
        T(col, left, val, right) -> {
          if (x < val)
            balance(T(col, insert_rec(left, x), val, right))
          else if (x > val)
            balance(T(col, left, val, insert_rec(right, x)))
          else
            T(col, left, x, right) # already here
        })
}
insert <- function(tree, x) {
  tree <- insert_rec(tree, x)
  tree$col <- B
  tree
}

balance <- function(tree) {
  cases(tree,
        T(B, T(R, a, x, T(R, b, y, c)), z, d) -> T(R, T(B,a,x,b), y, T(B,c,z,d)),
        T(B, T(R, T(R, a, x, b), y, c), z, d) -> T(R, T(B,a,x,b), y, T(B,c,z,d)),
        T(B, a, x, T(R, b, y, T(R, c, z, d))) -> T(R, T(B,a,x,b), y, T(B,c,z,d)),
        T(B, a, x, T(R, T(R, b, y, c), z, d)) -> T(R, T(B,a,x,b), y, T(B,c,z,d)),
        otherwise -> tree)
}

tree <- E
for (i in sample(2:4))
  tree <- insert(tree, i)
for (i in 1:6) {
  cat(i, " : ", member(tree, i), "\n")
}
```
