---
title: "Huffman coding"
author: "Thomas Mailund"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Huffman coding}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

[Huffman coding](https://en.wikipedia.org/wiki/Huffman_coding) is a compression approach based on character frequencies. To code a string, we work out the frequency of each letter in the string and then build a tree where we put the letters in the leaves and structure the tree such that the most frequent letters are closest to the root. We then encode each letter by the path we must take in that tree to get to the leaf that contain that letter. We need one bit for each edge we follow; zero, say, if we go left and one if we go right.

In [Haskell: the Craft of Functional Programming](https://amzn.to/2GHf6b6), there is an example of implementing Huffman coding in Haskell. In this vignette, I will adapt that to R with `pmatch` patterns.

## Data types

```haskell
data Tree = Leaf Char Int | Node Int Tree Tree
data Bit = L | R deriving (Eq, Show)
type HCode = [ Bit ]
type Table = [ (Char, HCode) ]
```

```{r}
library(pmatch)

Tree := Leaf(char : character, count : integer) | 
        Node(count : integer, left : Tree, right : Tree)
Bit := L | R
```

```{r}
llist := NIL | CONS(car, cdr : llist)
```

```{r}
pair := PAIR(x, y)
```

## Computing frequencies

```haskell
mergeSort :: ([a] -> [a] -> [a]) -> [a] -> [a]
mergeSort merge xs
  | length xs < 2 = xs
  | otherwise
      = merge (mergeSort merge first) (mergeSort merge second)
        where
          first = take half xs
          second = drop half xs
          half = (length xs) `div` 2
```

```{r}
llength <- function(llist, acc = 0) {
    cases(
        llist,
        NIL -> acc,
        CONS(car, cdr) -> llength(cdr, acc + 1)
    )
}

llrev <- function(llist, acc = NIL) {
    pmatch::cases(
        llist,
        NIL -> acc,
        CONS(car, cdr) -> llrev(cdr, CONS(car, acc))
    )
}

lltake <- function(llist, k, acc = NIL) {
    if (k == 0) return(llrev(acc))
    cases(
        llist,
        NIL -> stop("There were less than k elements in the list"),
        CONS(car, cdr) -> lltake(cdr, k - 1, CONS(car, acc))
    )
}
lldrop <- function(llist, k, acc = NIL) {
    if (k == 0) return(llist)
    cases(
        llist,
        NIL -> stop("There were less than k elements in the list"),
        CONS(car, cdr) -> lldrop(cdr, k - 1)
    )
}
```

```{r}
merge_sort <- function(merge, xs) {
    n <- llength(xs)
    if (n < 2) return(xs)
    half <- n %/% 2 
    first <- lltake(xs, half)
    second <- lldrop(xs, half)
    merge(merge_sort(merge, first), merge_sort(merge, second))
}
```

```haskell
alphaMerge :: [(Char,Int)] -> [(Char,Int)] -> [(Char,Int)]
alphaMerge xs [] = xs
alphaMerge [] ys = ys
alphaMerge ((p,n) : xs) ((q,m) : ys)
  | (p == q)  = (p,n+m) : alphaMerge xs ys
  | (p < q)   = (p,n) : alphaMerge xs ((q,m) : ys)
  | otherwise = (q,m) : alphaMerge ((p,n) : xs) ys
```

### Sorting alphabetically and counting characters

```{r}
alpha_merge <- function(xs, ys) {
    cases(
        ..(xs, ys),
        ..(xs, NIL) -> xs,
        ..(NIL, ys) -> ys,
        ..(CONS(PAIR(p,n), xs), CONS(PAIR(q,m), ys)) -> {
            if (p == q) {
                CONS(PAIR(p,n+m), alpha_merge(xs, ys))
            } else if (p < q) {
                CONS(PAIR(p,n), alpha_merge(xs, CONS(PAIR(q,m), ys)))
            } else {
                CONS(PAIR(q,m), alpha_merge(CONS(PAIR(p,n), xs), ys))
            }
        }
    )
}
```

```{r}
llist_from_list <- function(x) {
    llist <- NIL
    n <- length(x)
    while (n > 0) {
        llist <- CONS(x[[n]], llist)
        n <- n - 1
    }
    llist
}
```

```{r}
xs <- llist_from_list(list(
    PAIR("a", 1), PAIR("b", 2)
))
ys <- llist_from_list(list(
    PAIR("a", 3), PAIR("c", 3)
))
alpha_merge(xs, ys)
```

```{r}
llconcat <- function(xs, ys) {
    cases(xs, 
          NIL -> ys,
          CONS(x,xs) -> CONS(x, llconcat(xs, ys)))
}
zs <- llconcat(xs, ys)
merge_sort(alpha_merge, zs)
merge_sort(alpha_merge, llrev(zs))
```

```haskell
freqMerge :: [(Char,Int)] -> [(Char,Int)] -> [(Char,Int)]
freqMerge xs [] = xs
freqMerge [] ys = ys
freqMerge ((p,n):xs) ((q,m):ys)
  | (n < m || (n == m && p < q)) = (p,n) : freqMerge xs ((q,m):ys)
  | otherwise                    = (q,m) : freqMerge((p,n):xs) ys
```

### Sorting based on character frequency

```{r}
freq_merge <- function(xs, ys) {
    cases(
        ..(xs, ys),
        ..(xs, NIL) -> xs,
        ..(NIL, ys) -> ys,
        ..(CONS(PAIR(p,n), xs), CONS(PAIR(q,m), ys)) -> {
            if (n < m || (n == m && p < q)) {
                CONS(PAIR(p,n), freq_merge(xs, CONS(PAIR(q,m), ys)))
            } else {
                CONS(PAIR(q,m), freq_merge(CONS(PAIR(p,n), xs), ys))
            }      
        }
    )
}
```

```{r}
xs <- llist_from_list(list(
    PAIR("a", 5), PAIR("b", 3), PAIR("c", 2)
))
merge_sort(freq_merge, xs)
```

### Computing frequencies

```haskell
frequency :: [Char] -> [(Char,Int)]
frequency
  = mergeSort freqMerge . mergeSort alphaMerge . map start
    where
      start ch = (ch, 1)
```

```{r}
llmap <- function(llist, f, acc = NIL) {
    pmatch::cases(
        llist,
        NIL -> llrev(acc),
        CONS(car, cdr) -> llmap(cdr, f, CONS(f(car), acc))
    )
}
```

```{r}
string_to_list <- function(x)
    llist_from_list(strsplit(x, "")[[1]])

string_to_list("foobar")
```

```{r}
library(magrittr)
frequency = . %>% 
    string_to_list() %>% 
    llmap(function(char) PAIR(char, 1L)) %>%
    merge_sort(alpha_merge, .) %>%
    merge_sort(freq_merge, .)

frequency("foo")
```

## Building trees

```haskell
makeTree :: [ (Char,Int) ] -> Tree
makeTree = makeCodes . toTreeList

toTreeList :: [ (Char,Int) ] -> [Tree]
toTreeList = map (uncurry Leaf)
```

```{r}
to_tree_list <- . %>% llmap(
    function(x) cases(x, PAIR(char,count) -> Leaf(char, count))
)
xs <- "foo" %>% string_to_list() %>% llmap(function(char) PAIR(char, 1L))
to_tree_list(xs)
```

```haskell
makeCodes :: [ Tree ] -> Tree
makeCodes [t] = t
makeCodes ts = makeCodes (amalgamate ts)
```

```{r}
make_codes <- function(ts) {
    cases(
        ts,
        CONS(t, NIL) -> t,
        ts -> make_codes(amalgamate(ts))
    )
}
```

```haskell
amalgamate :: [ Tree ] -> [ Tree ]
amalgamate (t1:t2:ts) = insTree (pair t1 t2) ts
```

```{r}
amalgamate <- function(ts) {
    cases(
        ts,
        CONS(t1, CONS(t2, ts)) -> ins_tree(pair(t1, t2), ts)
    )
}
```

```haskell
pair :: Tree -> Tree -> Tree
pair t1 t2 = Node (v1 + v2) t1 t2
             where v1 = value t1
                   v2 = value t2

value :: Tree -> Int
value (Leaf _ n)   = n
value (Node n _ _) = n
```

```{r}
value <- function(tree) {
    cases(
        tree,
        Leaf(., n) -> n,
        Node(n, ., .) -> n
    )
}
pair <- function(t1, t2) Node(value(t1) + value(t2), t1, t2)
```

```haskell
insTree :: Tree -> [ Tree ] -> [ Tree ]
insTree t1 []          = [t1]
insTree t1 (t2 : rest) =
  if v1 < v2 then t1 : t2 : rest else t2 : insTree t1 rest
  where v1 = value t1
        v2 = value t2
```

```{r}
ins_tree <- function(t, ts) {
    cases(
        ..(t, ts),
        ..(t, NIL) -> CONS(t, NIL),
        ..(t1, CONS(t2, rest)) -> {
            v1 <- value(t1)
            v2 <- value(t2)
            if (v1 < v2) {
                CONS(t1, CONS(t2, rest))
            } else {
                CONS(t2, ins_tree(t1, rest))
            }
        }
    )
}
```

```haskell
makeTree :: [ (Char,Int) ] -> Tree
makeTree = makeCodes . toTreeList
```

```{r}
make_tree <- . %>%
    to_tree_list() %>%
    make_codes()
```

```{r}
"foo" %>% frequency() %>% make_tree()
```
